What does it take to create with domain-appropriate tools?

### 2020-05-14 convivial-2020/manuscript.tex
More amendments

\title{What does it take to create with domain-appropriate tools?}
\subtitle{A case study on the ``OROM'' system}
\author[ ]{Joel Jakubovic}
\affiliation[ ]{University of Kent, Canterbury}

@@

\newcommand{\joel}[1]{}
\newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
\newcommand{\OSFA}{One-Size-Fits-All }

\begin{abstract}
There is a One-Size-Fits-All quality to languages, APIs and even programming itself. Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code. This is a source of artificial difficulty in creating, understanding, and modifying software systems. No matter the domain, the author's design needs encoding into a form that does not resemble it.

@@

Amdahl's Law generalises the following idea: even when you spend hours
of effort doubling the performance of a component used 1\% of the time,
your reward is a system overall improved by a mere 0.5\%. Now, text
coding is certainly ubiquitous, the 99\% case in programming, so you
might wonder where I'm going with this. Well, a small improvement to
text editing, if adopted by everyone, certainly does have a massive
\emph{intermediate} effect --- but this only \emph{matters} to the
extent that text was helping the programmers in the first place. If my
goal is to draw or animate pictures, or create a digital synth from a
frequency spectrogram, then giving me the ability to auto-indent my SVG
markup is rather underwhelming as a productivity increase, as it doesn't
target the core of the enterprise that makes it so hard.

My experience of coding, most of the projects requiring shapes (such as
GUIs), leads me to conclude that no matter how much I improve my skill
at a particular language, knowledge of libraries or even general coding
ability, my predicament stays the same. Our basic method of creating
software is optimised for an ever-diminishing proportion of the software
we actually want to make; ill-optimised for the graphics, layout,
interactivity and and basic physics --- more on this later --- that we
usually require.

Whenever I work on these I feel stuck in a box I know I can never escape
from: that box is the text editor, a fixed conduit through which all
\emph{fundamental} changes to my program must pass. It's not a part of
the system I am building, so I can't even make use of features of the
thing I'm developing, to make its own development easier.

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with many
failed historical attempts to do better --- or at least, ones that
failed to catch on. And even worse than this is that in my very
\emph{language} here I am making the same mistake as the text editor ---
speaking in unqualified terms of ``better'' and ``worse'' as if there
really is a One-Size-Fits-All solution to software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best} in their particular context
--- to equip them to feasibly create the tools that suit them for the
thing they want to make. And second-order tools that suit them for
making the first-order tools, and so on. It would do no good to replace
text-imperialism with anything-else-imperialism, which is one
interpretation of calls for alternatives.

This dream goes beyond the familiar sense of what constitutes a
``craft'', as far as a strong melding of tool and material. Parallels
can be drawn with industrialisation and a strong division of labour: the
community as a whole produces its higher-order tools, but currently no
single person can have the same autonomy. A (future) software craft
could be expected to give this power to \emph{individuals}, instead of
the community alone. Whenever there are many small specialities
(e.g.~languages, tools, or subject areas) each serving many clients, the
One-Size-Fits-All style is the best one can hope for. Adaptation to
individual preferences and idiosyncrasies is only feasible when those
individuals can do it themselves.

What we need is some system that not only lets us create software in a
way that is ``close to the problem domain'' as decided by the
user-developer, but also can augment or change itself to adapt to a
different ``way of creating''. Existing systems seem to only have one of
these properties without the other: Smalltalk and LISP try to minimise
arbitrary commitments of language \emph{semantics} to this end, but
their being textual languages is a fairly tough commitment to break out
of. And it is not so hard to make a specific, \emph{hard-baked} visual
or alternative programming tool --- but it is hard to make it
re-programmable \emph{without} having to go back to \emph{its} textual
source code.

If someone wants to type out pictures in ASCII, let them --- whether
they do it for a challenge, or even if they find that more natural for
themselves. But equally, if I want to do it another way, then please
give me that affordance. This is how I segue into the software artefact
for which I have been attempting to build a natural representation.
True, it is a programmer's artefact, but it is still representative of
what any normal person has to do, insofar as:

@@

\hypertarget{the-orom-system}{%
\section{The OROM system}\label{the-orom-system}}

When I first read the paper ``Open, Reusable Object Models''
\cite{OROM}, I was hooked on its idea of a small but expressive starting
system that could be self-improved into anything. It describes a
late-bound\footnote{Fewer commitments; more things determined by runtime
  conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
  instances and messaging (methods) as in a distributed system; less
  emphasis on class hierarchies implementing traditional data structures}
objects and messaging environment that the authors call ``Id'' --- but I
shall refer to it as OROM, for pronunciation and Googleability.

Essentially, an OROM object is a block of state which can change as a
result of messages received by it. A message is sent by first
\emph{bind}-ing its name to its \emph{implementation}: specific code,
which is then run in the context of the receiver. ``Bind'' is
accomplished via another message send, this time to the receiver's
\emph{vtable}, or ``behaviour''. This is another object whose state
defines a particular way of associating ``message name'' to
``implementation code''. This continues up the vtable-chain, and
terminates at a base case. The system as a whole is ``bootstrapped''
into existence by its initialisation code. Each step of this code makes
use of any parts of the system set up by the previous steps.

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams. For example, the first acts of the
running system boil down to initialising the three or so objects. This
consists of allocating memory, interpreting it as a C struct and then
filling in fields in a mundane manner. I had great difficulty following
the specifics in my head, but when I drew tables in the style of their
diagrams I readily saw what what going on. I personally would have
preferred these to have been in the paper in the first place, but I am
not necessarily representative of those who read it. The
One-Size-Fits-All approach is perhaps unavoidable for static media like
print.

Other areas like messaging semantics could still be quite confusing.
After all, should we expect to be able to predict the entire future
evolution of a dynamical system from a static description of its initial
state, such as source code? Is this not why debuggers exist? Having
``de-compiled'' English text and C source code into object diagrams, it
is a shame to have to compile it all back to struct member assignments.

Worst of all, the reference system does not even have text I/O when it
is run, let alone some sort of GUI. That is, the (un)intended user
interface for this project is a C debugger! Faced with the necessity of
adding \emph{some} UI, it seemed a waste of effort to end up with a
system that must be continually polled for its current state at a
terminal prompt. If I naturally think of this system as 2D tables, why
can't that be how the running system looks? I do not have to keep
polling my eyes for what state my diagrams are in.

But further than that --- why can't the system be \emph{built} out of
tables in the first place? Shouldn't this be the main takeaway from the
amount of time we spend prototyping, explaining and designing software
as diagrams on paper? Why must the ``natural representation'' be
restricted to the finished product?

@@

\hypertarget{orom-as-html-tables}{%
\subsection{OROM as HTML tables}\label{orom-as-html-tables}}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-html.png}
  \caption{OROM/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of OROM objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}. Figure
\ref{fig:orom-html} shows the OROM/HTML implementation \cite{orom-html},
in which they take the form of HTML tables.

Here, I was grateful for the browser's management of graphical layout,
resizable text fields, and keeping the DOM tree synchronised with what
one sees. This last property enabled me to make the decision to
\emph{directly} encode much of the system state in the DOM, achieving
basic liveness\footnote{``The thing on the screen is the actual thing''}
for the keys and values of obj-dicts.

@@

A key aspect of the OROM system is that there is an object \emph{graph}.
That is, obj-dicts can have entries pointing to other obj-dicts, without
restriction to a tree structure. Drawing arrows to denote this is a
no-brainer (and easy on paper), so I wanted it in my substrate for OROM.

A small problem to surmount first: even though I could hijack the
\texttt{\textless{}table\textgreater{}} for its display properties, what
element could I hijack to make arrows between arbitrary points? Luckily,
there were Scalable Vector Graphics (SVG) at my disposal, which could be
persuaded to display \texttt{\textless{}line\textgreater{}}s over the
\texttt{\textless{}div\textgreater{}}s. But another key feature of my
intended substrate was to be able to rearrange and resize the boxes. So
I would also need to detect changes to the position and size of an
element.

@@

\hypertarget{orom-as-svg-trees}{%
\subsection{OROM as SVG trees}\label{orom-as-svg-trees}}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg.png}
  \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In OROM/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
elements, reminiscent of diSessa's Boxer\cite{boxer}. This was a
significant departure from the table representation, and even though SVG
supports (some) nested HTML via
\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
the possibility of multiple levels of nesting.

Still, this version was far more challenging and took much longer to
reach a satisfactory state. Yet it is precisely this drudgery that
brings me to a better understanding of this paper's question: \emph{what
has it taken?} I shall discuss this in the form of broad patterns or
themes that stand out to me from my development experience, hammered
home by these OROM projects.

@@

In this section, I will present these and other seemingly
\emph{inevitable} demands of normal software, exemplified by OROM/SVG. I
will comment on how our programming platforms measure up to the task,
including my chosen platform of JavaScript and Web technologies.

@@

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor --- just one with no GUI.

@@

All humans learn a basic set of expectations about the things they see
around them. Some of these, such as ``things fall down'', are not
generally appropriate to software UIs --- perhaps because the screen has
a role in our lives a more of a table work surface rather than a
vertical wall, even if it is vertical in real life. The level of physics
in software tends to not involve force, or mass, or very much at all,
merely position and space; we could call it ``geometric physics''.

@@

\begin{itemize}
\tightlist
\item
  This number range \texttt{0000}--\texttt{FFFF} is like a space,
  addresses = points
\item
  Every point has at most one owner block
\item
  These blocks are contiguous, finite ranges (i.e.~1D boxes)
\end{itemize}

Hence the boilerplate involved to realise this in any domain with
something resembling space. Far from being a niche topic in games and
graphics, spatial partitioning algorithms and data structures have
surprising relevance to more ordinary software. Both memory allocation
and graphical layout are essential to today's; shame that only one of
those has been recognised as such --- and entered the runtime of modern
languages.

@@

In OROM, the obvious candidate for this is the obj-dicts, plus all
nested boxes in OROM/SVG. If I move the top-level rect, then I expect
its children to move with it. This is simply the translational physics
of \textbf{rigid bodies}: ``this set of points all move together''. Of
course, proper rigid bodies might also rotate and have mass, but this is
usually undesirable for UI elements.

Translational rigidity can be expressed as the points X and Y always
having the same displacement from each other. Or, when one point is
moved, the rest also move by the same delta. This is a problem of
\textbf{preserving the relationship over time}, which was a significant
area of OROM/SVG.

@@

If a platform does not provide a means to causally link and unlink bits
of live state, then this must form part of the standard boilerplate.
Such was the case in OROM/SVG. I implemented the system in an OOP
fashion, and the Observable class is the most widely used. It wraps a
current value and a list of subscribers, notifying them when it changes.

Getting an object to follow the mouse pointer (e.g.~when dragging) is
\emph{conceptually} very simple: an ``always equal'' relation. In
OROM/SVG founded on live-state, this can be expressed in much the same
way:

@@

We are conditioned to only think of these in their most general forms.
But the ``vector graphics'' I use in OROM/SVG are just rects, lines,
circles and text; a fraction of the full capability of SVG. The
``geometric physics'' I use is dwarfed by fully general 2D or 3D physics
engines. The only layout algorithm I had the patience to implement was a
simple way to expand a list of boxes to fit in a new child at the
bottom. The affordance to place and size boxes \emph{manually} is a
convenient substitute, when required infrequently. Yet search for
material on layout algorithms, and it can seem like Fully General Linear
Inequality Solvers like Cassowary \cite{cassowary} are all there is.

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such generality might make it
\emph{more} cumbersome to express what I wanted in OROM. As the old
wisdom goes, there's no point expressing a simple regex search as an
arbitrary Turing machine; my boxes don't \emph{have} a moment of inertia
or a mass and I don't \emph{need} a linear optimisation solver for my
space management --- for the time being. Under the theme of
domain-appropriate tools, I think it is worth designing interfaces for
\emph{smaller-scale} instances of these areas and exploring what they
can express.

@@

When the \emph{normal} problem domain ``requires batteries'', yet the
tools available say ``batteries not included'', it is quite reasonable
to want the batteries to be part of the platform. This section details
the ``batteries not included'' I found myself polyfilling in OROM/SVG.

@@

One can do this to all objects, even those of Web APIs. For example, I
use it to connect SVG nodes to the OROM system: when a
\texttt{\textless{}rect\textgreater{}} is created, it is ``annotated''
with a \texttt{userData} pointer to the box that it is part of. Then,
when clicked, the box can be found immediately.

This could be described as the box ``wrapping'' the
\texttt{\textless{}rect\textgreater{}}, but risks impliying that the
\texttt{\textless{}rect\textgreater{}} merely needs to \emph{live
inside}, or \emph{be referenced by}, the box. In this case, the
\emph{reverse} association is also needed, because clicks can only enter
the system \emph{through} SVG in the first place. What is significant
here is that JavaScript lets us add such a link to an entity that
\emph{we did not design}.

This term, ``user data'', often crops up in libraries with the same
idea: to associate arbitrary data with an API object. But in languages
that do not permit such ``annotation'', this has to be explicitly
\emph{designed in} ahead of time. If it wasn't, then it has to be done
in a roundabout way --- for example, a separate \texttt{userData} lookup
table, indexed by the memory address of the API object.

@@

The simple desire to move and resize boxes with the mouse motivated a
lot of the concepts in Sections
\ref{basic-assumptions-about-physical-objects} and
\ref{maintaining-relationships-over-time}. This problem could be
considered a microcosm of OROM/SVG: what's a natural way I
\emph{conceive} of this behaviour, and could I implement it that way?

@@

I like to see what I'm doing, so I wanted these rods to be visible and
thus somehow \emph{present in the SVG}. This was not hard; in SVG I draw
a \texttt{\textless{}line\textgreater{}}, though the background
machinery of the Rod class does all the work. There is also a Point
class that is used wherever manipulable points (SVG circles) are
intended.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can show this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure}

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy\footnote{This highlights the mismatch between the
  tree-based DOM and any system that is graph-structured.}, but it is
still useful. However, it conflicts with my early decision to have Point
objects all share the global co-ordinate system (this was to ensure that
simple relations, such as a point following the mouse pointer, are not
infuriating to express.) Still, it was necessary in the case of certain
elements --- especially those which must transcend the tree structure
altogether, like arrows between boxes --- to bite this bullet, one way
or another.

@@

Each API has its own conventions, including a way of naming and
structuring expressions --- an ontology \cite{crit-semprola}. The
One-Size-Fits-All approach is exemplified in such interfaces. For
example, in SVG we express a rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height --- and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

>>>>

\newcommand{\OROM}{Id}
\title{What does it take to create with domain-appropriate tools?}
\subtitle{Reflections on implementing the ``Id{}'' system}
\author[ ]{Joel Jakubovic}
\affiliation[ ]{University of Kent, Canterbury}

@@

\newcommand{\joel}[1]{}
\newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
\newcommand{\OSFA}{One-Size-Fits-All}
\newcommand{\xywh}{\texttt{x},\texttt{y},\texttt{width},\texttt{height}}

\begin{abstract}
There is a One-Size-Fits-All{} quality to languages, APIs and even programming itself. Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code. This is a source of artificial difficulty in creating, understanding, and modifying software systems. No matter the domain, the author's design needs encoding into a form that does not resemble it.

@@

Amdahl's Law generalises the following idea: even when you spend hours
of effort doubling the performance of a component used 1\% of the time,
your reward is a system overall improved by a mere 0.5\%. Now, text
coding is certainly ubiquitous, the 99\% case in programming. A small
improvement to text editing, if adopted by everyone, certainly does have
a massive \emph{intermediate} effect---but this only \emph{matters} to
the extent that text was helping us in the first place. If my goal is to
draw or animate pictures, or create a digital synth from a frequency
spectrogram, then giving me the ability to auto-indent my SVG markup is
rather underwhelming as a productivity increase, as it doesn't target
the core of the enterprise that makes it so hard.

As a programmer, I often feel stuck in a box I know I can never escape
from: that box is the text editor, a fixed conduit through which all
\emph{fundamental} changes to my program must pass. It's not a part of
the system I am building, so I can't even make use of features of the
thing I'm developing, to make its own development easier.

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with all
sorts of concrete systems that failed to catch on. Even worse than this,
is that in my very \emph{language} here I am making the same mistake as
the text editor---speaking in unqualified terms of ``better'' and
``worse'' as if there really is a One-Size-Fits-All{} solution to
software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best}\footnote{To be clear: if
  someone \emph{wants} to type out pictures in ASCII, let them---whether
  they do it for a challenge, or even if they find that more natural for
  themselves. But equally, if I want to do it another way, I should have
  that affordance.} in their particular context---to equip them to
feasibly create the tools that suit them for the thing they want to
make. And second-order tools that suit them for making the first-order
tools, and so on. It would do no good to replace text-imperialism with
anything-\emph{else}-imperialism, which is one interpretation of calls
for alternatives.

This dream goes beyond the familiar sense of what constitutes a
``craft'', as far as a strong melding of tool and material. Parallels
can be drawn with industrialisation and a strong division of labour: the
community as a whole produces its higher-order tools, but currently no
single person can have the same autonomy. A (future) software craft
could be expected to give this power to \emph{individuals}, instead of
the community alone. Whenever there are many small specialities
(e.g.~languages, tools, or subject areas) each serving many clients, the
One-Size-Fits-All{} style is the best one can hope for. Adaptation to
individual preferences and idiosyncrasies is only feasible when those
individuals can do it themselves.

What we need is some system that not only lets us create software in a
way that is ``close to the problem domain'' as decided by the
user-developer, but also can augment or change itself to adapt to a
different ``way of creating''. Existing systems seem to only have one of
these properties without the other: Smalltalk and LISP try to minimise
arbitrary commitments of language \emph{semantics} to this end, but
their being textual languages is a fairly tough commitment to break out
of. And it is not so hard to make a specific, \emph{hard-baked} visual
or alternative programming tool---but it is hard to make it
re-programmable \emph{without} having to go back to \emph{its} textual
source code.

All these considerations attract me to the design of a particular system
called ``Id'', as a potential way out of this mess \cite{OROM}. True, it
is a programmer's artefact, but it is still representative of what any
normal person has to do, insofar as:

@@

What follows in this paper is an account of how I went about building
Id, atop a preliminary ``box'' substrate.
Section\textasciitilde{}\ref{the-system} gives an overview of Id.
Section\textasciitilde{}\ref{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}
discusses the work involved in obtaining said domain-appropriate
substrate, and Section\textasciitilde{}\ref{patterns-and-polyfilling}
continues with an interpretation of the labour costs of doing this. In
Section\textasciitilde{}\ref{the-system-as-a-part-of-the-solution}, we
return to the Id system and its role in supporting flexible software.
Section\textasciitilde{}\ref{conclusions-and-future-work} addresses the
question in the title, and concludes with next steps for the project.

\hypertarget{the-system}{%
\section{\texorpdfstring{The Id{}
system}{The  system}}\label{the-system}}

When I first read the paper ``Open, Reusable Object Models''
\cite{OROM}, I was hooked on its idea of a small but expressive starting
system that could be self-improved into anything. It describes a
late-bound\footnote{Fewer commitments; more things determined by runtime
  conditions}, Smalltalk-style\footnote{OOP with more emphasis on object
  instances and messaging, as in a distributed system; less emphasis on
  class hierarchies implementing traditional data structures} objects
and messaging environment that the authors call ``Id''.

An Id{} \emph{object} is a block of state which can change as a result
of messages received by it. Messaging (analogous to \emph{method
invocation} in mainstream OOP) works as follows:

\begin{itemize}
\tightlist
\item
  A message is sent by first \emph{bind}-ing its name to its
  \emph{implementation}: specific code, which is then run in the context
  of the receiver \(R\).
\item
  This ``bind'' step is accomplished by sending a further message; this
  time, to the receiver's \emph{vtable} \(V(R)\). A vtable is another
  object that maps ``message name'' to ``implementation code''---it's
  analogous to a ``class'' in mainstream OOP.
\item
  This initial ``bind'' message triggers a similar ``bind'' to
  \emph{its} vtable \(V(V(R))\), and so on: recursing up the
  vtable-chain, and terminating at a base case.
\item
  The higher levels of the vtable-chain mean that different kinds of
  vtables can be supported\footnote{As well as different kinds of
    ``kinds of vtables'', and so on.}, each implementing the ``bind''
  operation in its own way.
\end{itemize}

Inkeeping with its aims of minimality and self-describability, Id as a
whole is ``bootstrapped'' into existence by its initialisation code.
Each step of this code makes use of any parts of the system set up by
the previous steps.

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams. For example, the first acts of the
running system boil down to initialising the three or so objects. This
consists of allocating memory, interpreting it as a C struct and then
filling in fields in a mundane manner. I had great difficulty following
the specifics in my head, but when I drew tables in the style of their
diagrams I readily saw what what going on. I personally would have
preferred these to have been in the paper in the first place, but I am
not necessarily representative of those who read it\footnote{Besides
  that, the One-Size-Fits-All{} approach is perhaps unavoidable for
  static print media.}.

At any rate, in order to solidify my understanding, there was no better
way than to run the thing and explore it. And yet, having already
``de-compiled'' English text and C source code into object diagrams, it
is a shame to have to compile it all back to struct member assignments.
Worse, the reference system does not even have text I/O when it is run,
let alone some sort of GUI\footnote{That is, the (un)intended user
  interface for Id ends up being a C debugger!}. Faced with the
necessity of adding \emph{some} UI, it seemed a waste of effort to end
up with a system that must be continually polled for its current state
at a terminal prompt.

If I naturally think of this system as 2D tables, why can't that be how
the running system looks? I do not have to keep polling my eyes for what
state my diagrams are in. But further than that---why can't the system
be \emph{built} out of tables in the first place? Shouldn't this be the
main takeaway from the amount of time we spend prototyping, explaining
and designing software as diagrams on paper? Why must the ``natural
representation'' be restricted to the finished product?

@@

\hypertarget{as-html-tables}{%
\subsection{\texorpdfstring{Id{} as HTML
tables}{ as HTML tables}}\label{as-html-tables}}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-html.png}
  \caption{Id{}/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of Id{} objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}. Figure
\ref{fig:orom-html} shows the Id{}/HTML implementation \cite{orom-html},
in which they take the form of HTML tables.

Here, I was grateful for the browser's management of graphical layout,
resizable text fields, and keeping the DOM tree synchronised with what
one sees. This last property enabled me to make the decision to
\emph{directly} encode much of the system state in the DOM, achieving
basic liveness (``The thing on the screen is the actual thing'') for the
keys and values of obj-dicts.

@@

A key aspect of the Id{} system is that there is an object \emph{graph}.
That is, obj-dicts can have entries pointing to other obj-dicts, without
restriction to a tree structure. Drawing arrows to denote this is a
no-brainer (and easy on paper), so I wanted it in my substrate for Id{}.

Luckily, Scalable Vector Graphics (SVG) was at my disposal, which could
be persuaded to display \texttt{\textless{}line\textgreater{}}s over the
\texttt{\textless{}div\textgreater{}}s. But another key feature of my
intended substrate was to be able to rearrange and resize the boxes. So
I would also need to detect changes to the position and size of an
element.

@@

\hypertarget{as-svg-trees}{%
\subsection{\texorpdfstring{Id{} as SVG
trees}{ as SVG trees}}\label{as-svg-trees}}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg.png}
  \caption{Id{}/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via \emph{real arrows}.\label{fig:orom-svg}}
\end{figure}

In Id{}/SVG, shown in Figure \ref{fig:orom-svg}, obj-dicts are encoded
as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
elements, reminiscent of diSessa's Boxer \cite{boxer}. This was a
significant departure from the table representation, and even though SVG
supports (some) nested HTML via
\texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
the possibility of multiple levels of nesting.

Still, this version was far more challenging and took much longer to
reach a satisfactory state. Yet it is precisely this drudgery that
brings me to a better understanding of this paper's question: \emph{what
has it taken?} I shall discuss this in the form of broad patterns or
themes that stand out to me from my development experience, hammered
home by these Id{} implementations.

@@

In this section, I will present these and other seemingly
\emph{inevitable} demands of normal software, exemplified by Id{}/SVG. I
will comment on how our programming platforms measure up to the task,
including my chosen platform of JavaScript and Web technologies.

@@

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor---just one with no GUI.

@@

All humans learn a basic set of expectations about the things they see
around them. Some of these, such as ``things fall down'', are not
generally appropriate to software UIs---perhaps because the screen has a
role in our lives a more of a table work surface rather than a vertical
wall, even if it is vertical in real life. The level of physics in
software tends to not involve force, or mass, or very much at all,
merely position and space; we could call it ``geometric physics''.

@@

\begin{itemize}
\tightlist
\item
  This number range \texttt{0000}--\texttt{FFFF} is like a space
  (addresses = points)
\item
  Every point has at most one owner block (no overlap)
\item
  These blocks are contiguous, finite ranges (1D boxes)
\end{itemize}

Far from being a niche topic in games and graphics, spatial partitioning
algorithms and data structures have surprising relevance to more
ordinary software. Both memory allocation and graphical layout are
essential to today's; shame that only one of those has been recognised
as such---and made part of the standard libraries of programming
languages.

@@

In Id{}, the obvious candidate for this is the obj-dicts, plus all
nested boxes in Id{}/SVG. If I move the top-level rect, then I expect
its children to move with it. This is simply the translational physics
of \textbf{rigid bodies}: ``this set of points all move together''. Of
course, proper rigid bodies might also rotate and have mass, but this is
usually undesirable for UI elements.

Translational rigidity can be expressed as the points X and Y always
having the same displacement from each other. Or, when one point is
moved, the rest also move by the same delta. This is a problem of
\textbf{preserving the relationship over time}, which was a significant
area of Id{}/SVG.

@@

If a platform does not provide a means to causally link and unlink bits
of live state, then this must form part of the standard boilerplate.
Such was the case in Id{}/SVG: the polyfilled ``Observable'' class is
the most widely used. It wraps a current value and a list of
subscribers, notifying them when it changes.

Getting an object to follow the mouse pointer (e.g.~when dragging) is
\emph{conceptually} very simple: an ``always equal'' relation. In
Id{}/SVG founded on live-state, this can be expressed in much the same
way:

@@

We are conditioned to only think of these in their most general forms.
But the ``vector graphics'' I use in Id{}/SVG are just rects, lines,
circles and text; a fraction of the full capability of SVG. The
``geometric physics'' I use is dwarfed by fully general 2D or 3D physics
engines. The only layout algorithm I had the patience to implement was a
simple way to expand a list of boxes to fit in a new child at the
bottom. The affordance to place and size boxes \emph{manually} is a
convenient substitute, when required infrequently. Yet search for
material on layout algorithms, and it can seem like Fully General Linear
Inequality Solvers like Cassowary \cite{cassowary} are all there is.

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such generality might make it
\emph{more} cumbersome to express what I wanted in Id{}. As the old
wisdom goes, there's no point expressing a simple regex search as an
arbitrary Turing machine; my boxes don't \emph{have} a moment of inertia
or a mass and I don't \emph{need} a linear optimisation solver for my
space management---for the time being. Under the theme of
domain-appropriate tools, I think it is worth designing interfaces for
\emph{smaller-scale} instances of these areas and exploring what they
are still capable of expressing.

@@

When the \emph{normal} problem domain ``requires batteries'', yet the
tools available say ``batteries not included'', it is quite reasonable
to want the batteries to be part of the platform.
Section\textasciitilde{}\ref{what-did-it-take-and-why} will expand on
this further, but here I will list the ``batteries not included'' I
found myself polyfilling in Id{}/SVG.

@@

One can do this to all objects, even those of Web APIs. For example, I
use it to connect SVG nodes to the Id{} system: when a
\texttt{\textless{}rect\textgreater{}} is created, it is ``annotated''
with a \texttt{userData} pointer to the box that it is part of. Then,
when clicked, the box can be found immediately.

This could be described as the box ``wrapping'' the
\texttt{\textless{}rect\textgreater{}}, but risks implying that the
\texttt{\textless{}rect\textgreater{}} merely needs to \emph{live
inside}, or \emph{be referenced by}, the box. In this case, the
\emph{reverse} association is also needed (i.e.~mutual pointers); clicks
can only enter the system \emph{through} SVG in the first place. What is
significant here is that JavaScript lets us add such a link to an entity
that \emph{we did not design}.

This term, ``user data'', often crops up in libraries with the same
idea: to associate arbitrary data with an API object. But in languages
that do not permit such \emph{unanticipated} ``annotation'', this has to
be explicitly \emph{designed in} ahead of time. If it wasn't, then it
has to be done in a roundabout way---for example, a separate
\texttt{userData} lookup table, indexed by the memory address of the API
object.

@@

The simple desire to move and resize boxes with the mouse motivated a
lot of the concepts in Sections
\ref{basic-assumptions-about-physical-objects} and
\ref{maintaining-relationships-over-time}. This problem could be
considered a microcosm of Id{}/SVG: what's a natural way I
\emph{conceive} of this behaviour, and could I implement it that way?

@@

I like to see what I'm doing, so I wanted these rods to be visible and
thus somehow \emph{present in the SVG}. For this, I have a Point class
that is used wherever manipulable points (SVG circles) are needed, and a
Rod class that draws a \texttt{\textless{}line\textgreater{}} and
transmits endpoint deltas across its length.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can reveal this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure}

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy\footnote{This highlights the mismatch between the
  tree-based DOM and any system that is graph-structured.}, but it is
still useful. However, it conflicts with my early decision to have Point
objects all share the global co-ordinate system (this was to ensure that
simple relations, such as a point following the mouse pointer, are not
infuriating to express.) Still, it was necessary in the case of certain
elements---especially those which must transcend the tree structure
altogether, like arrows between boxes---to bite this bullet, one way or
another.

@@

Each API has its own conventions, including a way of naming and
structuring expressions---an ontology \cite{crit-semprola}. The
One-Size-Fits-All{} approach is exemplified in such interfaces. For
example, in SVG we express a rectangle by
\texttt{\textless{}rect x="10" y="10" width="600" height="400"\textgreater{}}.
The SVG specification \emph{defines} to the user that a rect simply
\textbf{is} a top-left corner, a width, and a height---and that's it.
However, this SVG-approved parametrisation of a rectangle is far from
the only one, and thus is, unsurprisingly, ill-fitted to some contexts.

For example, I find it natural to resize boxes by dragging any of their
four corners, so I wanted this in Id{}/SVG. In this context, a
``rectangle'' is \emph{seen as} four points: top-left, top-right,
bottom-left, bottom-right. Obviously this is not a \emph{minimal}
description, since given e.g.~the top-left and bottom-right, the other
two points can be inferred. But one way or another, to be able to drag
any of them, all four points must be present at \emph{some} level. This
alternative ontology was polyfilled in the form of a ``rect controls''
class that can be attached manually to any SVG
\texttt{\textless{}rect\textgreater{}}. The \texttt{x} and \texttt{y}
are subscribed to the top-left Point\footnote{In the usual case where
  the \texttt{\textless{}rect\textgreater{}} is part of a box, the
  top-left instead controls the parent
  \texttt{\textless{}g\textgreater{}}'s \texttt{transform}---but the
  idea is the same.}; \texttt{width} and \texttt{height} are subscribed
to Rod lengths.

Another example is to be found in the DOM's event listener model.
Conceptually, many ``events'' are in fact changes to the state of some
physical device. And both keyboard keys and mouse buttons, for example,
have two states---pressed, not pressed---which ought to make them
\emph{interchangeable} to some extent. Indeed, this is why there is such
a thing as ``key mapping''\footnote{For example, the player can
  re-assign the action ``shoot'' from its default mouse button to a
  keyboard key, or another action from keyboard to mouse, as it suits
  them.} in PC games. But the official ontology of the Web makes it
nontrivial to do this.
